Visualization and analysis of probability distributions of large temporal data

### 2022-02-28 report/response.Rmd

---
title: Response document
author: Sayani Gupta
output:
  MonashEBSTemplates::memo:
    includes:
      in_header: preamble2.tex
fontfamily: mathpazo
---

I thank my examiners for such thorough and constructive comments. The point by point description of changes are below: the examiners' comments are in red and my response is in black.

@@

Not changed as figure is already published. Furthermore, the figure already occupies one page.

## Chapter 3

> Q1. On page 33 there is a mention of the hakear package, without a reference.

Reference added as link on page 35. The package reference is also mentioned in the Supplementary section of the paper.

<!-- No change. -->
<!-- It appears in the abstract of the paper. For the thesis, I have added it since the abstract acts as a summary to the chapter. The package reference is mentioned in the Supplementary section of the paper. Moreover, the plan is to bring all the functions from `gravitas` to `hakear` when we publish it. -->

@@

> Q2. In the Introduction state the methods are for continuous variables.

Added "for a continuous univariate dependent variable" in the last paragraph of section 3.1 (Introduction) on page 39.

@@

> Q3. In Figure 3.1, the points are a bit hard to see, so it looks like whiskers extend beyond the data.

The following are changed in Figure 3.1 to address the comment on page 36:

- changed to theme_bw() instead of default
- outliers and jitters are made more transparent
- jitter width reduced to make it narrower

<!-- outlier.alpha (0.5 ->0.2) and geom_jitter(alpha = 0.04 -> 0.03). Also jitter-width decreased to 0.2. -->

@@

> Q4. On page 35, refer to months consistently using short or long names.

The comment is addressed by changing the x-axis labels to month-of-year for Figures 3.1 (page 36) and 3.2 (page 38).

@@

> Q5. On page 35 v refers to the number of variables, in Section 3.3.3 it refers to the variable.

On p38, $v$ changed to $p$, where $p$ represents the number of variables. This is analogous to many time granularities and their pairwise combinations and $v$ represents the univariate measured variable for which distributions are constructed for different time granularities and their pairs (added to Table 3.1 on page 42).

@@

> Q6. In the Figure 3.2 caption, I do not understand the comment beginning “Difference between the 90th. . . ”.
Also for the sentence “Energy consumption for (a)..”, it would clarify to insert the word “median”.

Caption changed on page 38 to address this.

<!-- clarification: If the difference between $90^{th}$ and $75^{th}$ percentile of energy distribution is low implies that energy usage is extremely -->

@@

Added text in the caption of Figure 3.3(a) and Section 3.2.2 (Notation) on page 41 to make it clear.

<!-- Clarification: Null distributions refer to the case when there is no significant distributional differences between categories of one or more granularities. Data in Fig 3.3 (a) is consistent with the null distribution. -->

> Q10. In table 3.1 $N_c$ is the number of cyclic granularities, whereas in the text (first sentence of 3.3.2 and
3.3.3) it is m.

Added $m$ to table 3.1 (page 42).
$m$: number of cyclic granularities to display together

@@

Added v:continuous univariate measured variable to the table on page 42.

> Q12. In the first sentence on the top of page 42, ordered and unordered are mixed up.

Done.

> Have you considered the setting where the facet variable levels are ordered?

@@

Caption changed on page 44 to address this comment.

<!-- Within-facet distances are illustrated in Panels 3) (when categories are un-ordered, shown only with respect to a1) and Panel 4) (when categories are ordered) -->

> Q14. In the pairwise distance measure on page 43, is there any adjustment made for varying numbers of observations across the levels of A and B? For example in Figure 3.3(d) if there are few values at xlevel=1 the comparison shown is less interesting

Added text in the section 3.2.3 under subsection "Characterizing distributions" on page 43 to clarify this.

@@

<!-- Clarification: There is no practical explanation for this, but a more empirical one - with smaller $n_x$ and $n_f$, the variance in the values of $wpd_{raw}$ is greater. As a result, our modeling technique that attempts to fit $wpd_{raw}$ to the number of comparisons ($n_x\times n_f$) performs poorly. The increased volatility in the values of $wpd_{raw}$ with a smaller number of levels is due to the fact that no asymptotic distribution for the statistic "maximum" (which is the foundation of $wpd_{raw}$) across such small number of values could be produced. -->

> Q17. In the 3.3.1 algorithm, there is $m$, and then $M$.

Changed on page 49.
<!-- 1.b) for $i \in {1, 2, \dots, M}$. -->

> Q18. In Table 3.2, I would suggest separating the $m = 1$ and $m = 2$ tables.

@@

$m=1$ removed from the main chapter and included in supplements.

> Q19. The presentation of the material on the simulation study on page 48 could be improved. Where is the notation $wpd_{l,s}$ used? Figure 3.5 shows the m = 2 results only. The text alludes to a simulation involving different underlying distributions, but this is not mentioned again.

Presentation of the simulation design and results corresponding to $m=2$ changed on page 50. Similar design and results for $m=1$, although important, are not included in the main chapter but in the supplements. Simulation results involving different underlying distributions are also presented in the supplementary paper.

@@

> Q21. In Figure 3.6 the axis tick labels should be smaller. The blue and orange are hard to see. The caption
should refer to the rug. Maybe show fewer panels?

Changed (page 52). Axis tick labels made smaller and the rug plots are removed to make the display less busy.

> Q22. In Figure 3.7 (a) the heatmaps need id labels. The grey color is missing from the legend. Maybe use a different colour in the heatmap for the significant comparisons. State the threshold for significance in the caption. In Table 3.3 the caption should explain the threshold. (Maybe use colour instead of stars to indicate significance?)

The following are changed in Figure 3.7 (a) (page 53).
- Different box borders (instead of id labels) are added such that (a) and (b) display the household ids in the same order as indicated by the same colour of the line plot in (b) and box border
- grey color explained in the text and caption

The threshold explained in caption in Table 3.3 (page 54).

@@

> Q1. While the author hints at other sources of similar data, I think it would be worthwhile to give the reader some further specific ideas here in the Introduction where similar complex data may occur, e.g., data from traffic sensors, movement data (turnstiles in metro stations or public buildings), and even more traditional measurements such as temperatures or precipitation over time that could benefit from the methods and software described in this thesis.

Added in the Introduction (page 2).

> Q2. It may also be helpful to state what is not be covered in this thesis: The spatial component of such data where an additional component could be latitude and longitude or some areal code, such as the post code or some administrative units. Extending the work from this thesis with spatial proximity methods for clustering (https://doi.org/10.1016/j.cageo.2011.12.017) [assuming that households in certain residential neighborhoods may have similar energy usage patterns] and trying spatial visualization methods (if possible) such as glyph maps (https:
//onlinelibrary.wiley.com/doi/abs/10.1002/env.2152) should be mentioned somewhere as a consideration for future work but that may become a future new thesis by itself.

Added in section 5.3.2 (Future ideas) on page 90.

<!-- The name of the thesis implies that only analyzing temporal data is covered. Extending it to spatial is too broad a scope. The limitations of the study cater to aspects that could be addressed with some more time. This is not the case for extending the spatial component, which, as accurately pointed out, would require time comparable to another thesis. -->

@@

Done.

> Q5. Before introducing a new distance measure in Section 3.2, I would like to see a literature review of existing distance measures and their limitations. Why is it necessary to introduce a new distance measure here? This is partially addressed on p. 41, but that should be placed earlier in the text. One clarifying question: Which density is represented by f in the equation? And shouldn't there be a dx at the end of the integral? Also, the distance measures that are mentioned at the end of that paragraph should be explained in one or two sentences.

Changed the following:

- Introduction (last but one paragraph on page 39) to incorporate comments on existing distance measures and their limitations.

@@

> Q16. Match the text on p. 50 and introduce the abbreviations from Figure 3.7 and Table 3.3, e.g., "hod" likely is "hour day" and so on. Introducing these abbreviations on p. 53 comes
too late.

Added the abbreviations on page 54 in "Choosing cyclic granularities of interest and removing clashes" under section 3.4.

> Q17. In Figure 3.7a, it is not clear what the eight heatmaps represent. Do they belong to id 1, ..., id 8 ? Some explanation in the text or listing the id information on the left would be helpful. Similarly, on p. 53, the text states: id 7 and 8 have the same significant harmonies. I do not know where to look at the heatmaps in Figure 3.7a to find this information (and what is mentioned under items 2. and 3. in the text). Some explanation and labeling is needed here.

Caption changed on page 53 to illustrate the plot better. Added box borders in (a) such that (a) and (b) display the household ids in the same order as indicated by the same color of the line plot in (b) and box border in (a). Added text on page 55 to state where to look at for item 2 and 3 in the text.

@@

> Q18. How were the harmony pairs selected and sorted in Figure 3.7 and Table 3.3? headmaps show 25 pairs, the parallel coordinate plot shows only 14 (same as Table 3.3). However, the last two are sorted differently, "hod wdwnd" (the first row in the table) is neither the first nor the last row in the parallel coordinate plot.

Changed the text description of Figure 3.7 on page 55 to address the comments.

Clarification 1: The harmony pairs are selected using the same procedure described in Chapter 2 (as mentioned on page 54). The harmonies in Table 3.3 are sorted according to id 1. The cyclic granularities in Figure 3.7 (a) is in a lexicographic order. The harmony pairs in the parallel coordinate plot (Figure 3.7 (c)) are arranged from highest to lowest values of $wpd$ averaged over all households and hence do not correspond to the first or last rows of Table 3.3.

Clarification 2: Each household is represented by 25 tiles, each tile representing a pair of cyclic granularities. The colors (in shades of red) represent the value of $wpd$ for each of the harmony pairs (in Table Table 3.3) and the grey tiles correspond to clashes. A darker shade of red corresponds to higher values of $wpd$.

@@

Caption rephrased to make it more clear on page 56.

<!-- For id: 7, there is more distributional difference between consecutive hours of the day compared to id:1. Also all the days of the week are different from each other in id:7. id:1 has similar behavior across all days within both weekdays and weekends. For this reason, the wpd for id:7 has a higher value than id:1 as the similar behavior within weekdays and weekends balances out the difference between weekdays and weekends. The display with hod vs wdwnd would have shown higher values of wpd for id:1 (as can be verified from Table 3). -->

> Q22. I didn't ask earlier, but how were the eight households for Section 3.4 being selected? Was this based on a random sample, or manually do obtain rather different energy usage patterns? This should be answered earlier on, but this also should be addressed in the Discussion: What can be expected when working with the full data set of Chapter 2, i.e., about 13,000. You indicate:
A future direction of work is to be able to explore and compare many individuals/subjects together for similar patterns across significant granularities. What would be computationally (and visually) feasible as 13,000 is a few orders of magnitude
higher than the number of eight households used in this example? Clearly, no full answer is expected here, but rather some speculation of what could possibly be done in the future (or what is done in the next chapter).

- Text added in the first paragraph of Section 3.4 to elaborate how households were selected (page 51).

@@

> Q5. The text on p. 68 speaks of Figure 4.3(b) and Figure 4.3 (right). Either add letters a and b or speak of left and right.

Done.

>Q6. Similar to Chapter 3, supplementary material should be specified in more detail, e.g., on p. 68, p. 70, p. 71, and p. 75.

Added links to supplementary material in all the relevant places.

### 2022-02-28 Rmd/03-hakear.Rmd
fine tuning all comments and cross checking with thesis all done

```{r id2-new2, fig.cap = "Distribution of energy consumption displayed through area quantile plots across two cyclic granularities month-of-year and hour-of-day and two households. The black line is the median, whereas the orange band covers the 25th to 75th percentile and the green band covers the 10th to 90th percentile. Difference between the 90th and 75th quantiles is less for (Jan, Feb) for the first household (a), suggesting that it is a more frequent user of air conditioners than the second household (b). Distribution of energy usage for (a) changes across both granularities, whereas for (b) daily pattern stays same irrespective of the months."}
```

@@

A common way to measure divergence between distributions is the Kullback-Leibler (KL) divergence [@Kullback1951-jy]. The KL divergence denoted by $D(q_1||q_2)$ is a non-symmetric measure of the difference between two probability distributions $q_1$ and $q_2$ and is interpreted as the amount of information lost when $q_2$ is used to approximate $q_1$. The KL divergence is not symmetric and hence can not be considered as a "distance" measure. The Jensen-Shannon divergence [@Menendez1997-in] based on the Kullback-Leibler divergence is symmetric and has a finite value. Hence, in this paper, the pairwise distances between the distributions of the measured variable are obtained through the square root of the Jensen-Shannon divergence, called Jensen-Shannon distance (JSD), and defined by
$$
  JSD(q_1||q_2) = \frac{1}{2}D(q_1||M) + \frac{1}{2}D(q_2||M),
$$
where $M = \frac{q_1+q_2}{2}$ and $D(q_1||q_2) := \int^\infty_{-\infty} q_1(x)log(\frac{q_1(x)}{q_2(x)})$ is the KL divergence between distributions $q_1$ and $q_2$. There are other ways to obtain distance between distributions like Hellinger distance, total variation distance and Fisher information metric (all of which are special cases of f-divergence), but Jensen-Shannon distance was chosen for ease of computation.

@@

The simulation results (in Section \ref{sec:simulations}) show that the distribution of $\wpdsub{glm}$ under the null design is similar for high $\nx$ and $\nf$ (levels higher than $5$) but less so for lower values of $\nx$ and $\nf$. An empirical explanation for this is that with smaller $n_x$ and $n_f$, the variance in the values of $wpd_{raw}$ is greater [Figure 13 of supplements](https://github.com/Sayani07/paper-hakear/blob/master/paper/paper-supplementary.pdf). As a result, our modeling approach that fits $wpd_{raw}$ to the number of comparisons ($n_x\times n_f$) performs poorly. In order to address this, a combination approach is proposed where we use a permutation approach for categories with small numbers of levels, and a modeling approach for categories with higher numbers of levels. This ensures that the computational load of the permutation approach is alleviated while maintaining a similar null distribution across different categories. This approach, however, requires that the adjusted variables from the two approaches are brought to the same scale. We define $\wpdsub{glm-scaled} = \wpdsub{glm}\times \sigma^2_{\text{perm}}/\sigma^2_{\text{glm}}$ as the transformed $\wpdsub{glm}$ with a similar scale as $\wpdsub{perm}$. The adjusted measure from the combination approach, denoted by $\wpd$ is then defined as follows:
\begin{equation}
\wpd = \begin{cases}
         \wpdsub{perm},       & \text{if $\nx, \nf \le 5$};\\
         \wpdsub{glm-scaled}  & \text{otherwise}.\\
       \end{cases}
\end{equation}

@@

Let $v_{i, t}$ denote the electricity demand for the $i^{th}$ household in time period $t$. The series $\{v_{i, 1},\dots,v_{i,T}\}$ is the linear granularity corresponding to half-hour since the interval of the `tsibble` is 30 minutes. We consider coarser linear granularities like hour, day, week and month from the commonly used Gregorian calendar. From the four linear granularities of hour, day, week, and month, we obtain $N_C = 4\times 3/2 = 6$ cyclic granularities:  "hour_day", "hour_week", "hour_month", "day_ week", "day_month" and "week_month" abbreviated as *hod*, *how*, *hom*, *dow*, *dom* and *wnwd* respectively. Further, we add cyclic granularity day-type ("wknd wday") to capture weekend and weekday behavior. Thus, seven cyclic granularities are considered to be of interest. The pairs of cyclic granularities ($C_{N_C}$) will have $7\times 6=42$ elements. The set of possible harmonies $H_{N_C}$ from $C_{N_C}$ are chosen by removing clashes using procedures described in @Gupta2021-gravitas. Table \ref{tab:tab-rank-8} shows $14$ harmony pairs that belong to $H_{N_C}$.

@@

We can now see some patterns that were not discernible in Panel (b), including:

@@

<!-- A rough estimate of working with $\wpd$ working with two granularities with $(l,m)$ levels: ~(30s - 300s for 6<=l,m<=50) Computing $\wpd$ is computationally more expensive for cyclic granularities with lower levels $\nx,\nf<=5$. -->
<!-- patterns are interesting and others are not. The subjects for which behaviors across a group of cyclic granularities are similar, could be grouped together as their periodic behavior is similar. to group subjects -->
The computation time of $\wpd$ for a cyclic granularity or harmony is determined by its levels. For example, in a harmony with $6 \leq\nx,\nf \leq 50$, the computation time grows with the number of levels and ranges from $30-300$s. $\wpd$ becomes computationally more expensive for cyclic granularities with lower levels $\nx,\nf \leq 5$. It is around $40-50$ minutes for $\nx,\nf = \{2,3\}$. Even with parallel computation, computing $\wpd$ for many individuals and cyclic granularity/harmonies takes a long time. One solution could be to obtain the significant granularities for a smaller number of individuals in order to get a sense of the significant granularities for the context and then computing $\wpd$ only for those granularities for other individuals. <!--For example, in the case of electricity data, the important granularities for most people were *wnwd*, *hod*, *dow*, and *dom*. After this fact has been established for a few people, $\wpd$ can be computed exclusively for these granularities to save time for others.--> However, information may be lost for individuals with different sets of significant granularities. In any case, additional studies may be conducted to speed up the processing of cyclic granularities with low levels.

>>>>

```{r id2-new2, fig.cap = "Distribution of energy consumption displayed through area quantile plots across two cyclic granularities month-of-year and hour-of-day and two households. The black line is the median, whereas the orange band covers the $25^{th}$ (lower) to $75^{th}$ (upper) percentile and the green band covers the $10^{th}$ (lower) to $90^{th}$ (upper) percentile. Difference between the $90^{th}$ and $75^{th}$ percentile is less for (Jan, Feb) for the first household (a), suggesting that it is a more frequent user of air conditioners than the second household (b). Distribution of energy usage for (a) changes across both granularities, whereas for (b) daily pattern stays same irrespective of the months."}
```

@@

A common way to measure divergence between distributions is the Kullback-Leibler (KL) divergence [@Kullback1951-jy]. The KL divergence denoted by $D(q_1||q_2)$ is a non-symmetric measure of the difference between two probability distributions $q_1$ and $q_2$ and is interpreted as the amount of information lost when $q_2$ is used to approximate $q_1$. The KL divergence is not symmetric and hence can not be considered as a "distance" measure. The Jensen-Shannon divergence [@Menendez1997-in] based on the Kullback-Leibler divergence is symmetric and has a finite value. Hence, in this paper, the pairwise distances between the distributions of the measured variable are obtained through the square root of the Jensen-Shannon divergence, called Jensen-Shannon distance (JSD), and defined by
$$
  JSD(q_1||q_2) = \frac{1}{2}D(q_1||M) + \frac{1}{2}D(q_2||M),
$$
where $M = \frac{q_1+q_2}{2}$ and $D(q_1||q_2) := \int^\infty_{-\infty} q_1(x)log(\frac{q_1(x)}{q_2(x)})dx$ is the KL divergence between distributions $q_1$ and $q_2$. There are other ways to obtain distance between distributions like Hellinger distance, total variation distance and Fisher information metric (all of which are special cases of f-divergence), but Jensen-Shannon distance was chosen for ease of computation.

@@

The simulation results (in Section \ref{sec:simulations}) show that the distribution of $\wpdsub{glm}$ under the null design is similar for high $\nx$ and $\nf$ (levels higher than $5$) but less so for lower values of $\nx$ and $\nf$. An empirical explanation for this is that with smaller $n_x$ and $n_f$, the variance in the values of $wpd_{raw}$ is greater [Figure 13 of supplements](https://github.com/Sayani07/paper-hakear/blob/master/paper/paper-supplementary.pdf). As a result, our modeling approach that fits $wpd_{raw}$ to the number of comparisons performs poorly. In order to address this, a combination approach is proposed where we use a permutation approach for categories with small numbers of levels, and a modeling approach for categories with higher numbers of levels. This ensures that the computational load of the permutation approach is alleviated while maintaining a similar null distribution across different categories. This approach, however, requires that the adjusted variables from the two approaches are brought to the same scale. We define $\wpdsub{glm-scaled} = \wpdsub{glm}\times \sigma^2_{\text{perm}}/\sigma^2_{\text{glm}}$ as the transformed $\wpdsub{glm}$ with a similar scale as $\wpdsub{perm}$. The adjusted measure from the combination approach, denoted by $\wpd$ is then defined as follows:
\begin{equation}
\wpd = \begin{cases}
         \wpdsub{perm},       & \text{if $\nx, \nf \le 5$};\\
         \wpdsub{glm-scaled}  & \text{otherwise}.\\
       \end{cases}
\end{equation}

@@

Let $v_{i, t}$ denote the electricity demand for the $i^{th}$ household in time period $t$. The series $\{v_{i, 1},\dots,v_{i,T}\}$ is the linear granularity corresponding to half-hour since the interval of the `tsibble` is 30 minutes. We consider coarser linear granularities like hour, day, week and month from the commonly used Gregorian calendar. From the four linear granularities of hour, day, week, and month, we obtain $N_C = 4\times 3/2 = 6$ cyclic granularities:  "hour_day", "hour_week", "hour_month", "day_ week", "day_month" and "week_month" abbreviated as *hod*, *how*, *hom*, *dow* and *dom*  respectively. Further, we add cyclic granularity day-type ("wknd wday") to capture weekend and weekday behavior (abbreviated as *wnwd*). Thus, seven cyclic granularities are considered to be of interest. The pairs of cyclic granularities ($C_{N_C}$) will have $7\times 6=42$ elements. The set of possible harmonies $H_{N_C}$ from $C_{N_C}$ are chosen by removing clashes using procedures described in @Gupta2021-gravitas. Table \ref{tab:tab-rank-8} shows $14$ harmony pairs that belong to $H_{N_C}$.

@@

We can now see some patterns in \ref{fig:dotplot-8}(a) that were not discernible in \ref{fig:dotplot-8}(b), including:

@@

<!-- A rough estimate of working with $\wpd$ working with two granularities with $(l,m)$ levels: ~(30s - 300s for 6<=l,m<=50) Computing $\wpd$ is computationally more expensive for cyclic granularities with lower levels $\nx,\nf<=5$. -->
<!-- patterns are interesting and others are not. The subjects for which behaviors across a group of cyclic granularities are similar, could be grouped together as their periodic behavior is similar. to group subjects -->
The computation time of $\wpd$ for a cyclic granularity or harmony is determined by its levels. For example, in a harmony with $6 \leq\nx,\nf \leq 50$, the computation time grows with the number of levels and ranges from $30$ to $300$ seconds. $\wpd$ becomes computationally more expensive for cyclic granularities with lower levels $\nx,\nf \leq 5$. It is around $40$ to $50$ minutes for $\nx,\nf = \{2,3\}$. To perform these computations, R version 4.0.1 (2020-06-06) with the platform: x86 64-apple-darwin17.0 (64-bit) running under: macOS Mojave 10.14.6 and [MonaRCH](https://docs.monarch.erc.monash.edu/) is utilized. Even with parallel computation, computing $\wpd$ for many individuals and cyclic granularity/harmonies takes a long time. One solution could be to obtain the significant granularities for a smaller number of individuals in order to get a sense of the significant granularities for the context and then computing $\wpd$ only for those granularities for other individuals. <!--For example, in the case of electricity data, the important granularities for most people were *wnwd*, *hod*, *dow*, and *dom*. After this fact has been established for a few people, $\wpd$ can be computed exclusively for these granularities to save time for others.--> However, information may be lost for individuals with different sets of significant granularities. In any case, additional studies may be conducted to speed up the processing of cyclic granularities with low levels.

### 2022-02-28 Rmd/04-gracsr.Rmd
fine tuning all comments and cross checking with thesis all done

The proposed methods could be extended when two granularities of interest interact and we want to group subjects based on the interaction of the two granularities. Consider a group that has a different weekday and weekend behavior in the summer but not in the winter. This type of combined behavior across granularities can be discovered by evaluating the distribution across combinations of categories for different interacting granularities (weekend/weekday and month-of-year in this example). As a result, in this scenario, we analyze a combination of categories generated from different distributions. Display of design and related results can be found in supplementary material.

@@

All of the approaches were fitted to each data design and to each combination of the considered parameters considered in Table \ref{tab:range-parameter}. The formed clusters have to match the design, be well separated, and have minimal intra-cluster variation. <!--It is possible to study these desired clustering traits visually in a more comprehensive way than just looking at index values.--> MDS and parallel coordinate graphs are used to demonstrate the findings, as well as an index value plot to provide direction on the number of clusters. <!--JS-based approaches corresponding to NQT and RS are referred to as JS-NQT and JS-RS respectively.--> In the following plots, results for JS-NQT are reported, and results with JS-RS or wpd-based distances are in the supplementary material.

@@

Clustering with the new distances is illustrated on the smart meter energy usage for a sample of customers from @smart-meter. The full data contains half-hourly general supply in kWh for 13,735 customers, resulting in 344,518,791 observations in total.<!--In most cases, electricity data is expected to have multiple seasonal patterns like daily, weekly or annual. We do not learn about these repetitive behaviors from the linear view because too many measurements all squeezed in that representation. Hence we transition into looking at cyclic granularities, that can potentially provide more insight on their repetitive behavior.--> The raw data for these consumers is of unequal length, with varying starting and end dates. Additionally, there were missing values in many series. (The supplementary material contains details from checking for systematic missingness.) Because our proposed methods evaluate probability distributions rather than raw data, these data issues are not problematic, unless there is any systematic structure related to granularities.

@@

It is common to filter data prior to fitting a supervised classification model using instance selection [@olvera2010review] which removes observations that might impede the model building. For clustering, this is analogous to identifying and removing nuisance observations. Prototype selection is more severe than instance selection, because only a handful of cases is selected. @Cutler1994-hm proposed a method called archetypal analysis which has inspired this approach but the procedure we have used follows @Fan2021-bq. First, dimension reduction such as t-SNE, MDS or PCA is used to project the data into a 2D space. Second, a few "anchor" customers far apart in 2D space are selected. Additional close neighbors to the anchors are selected. To check the selections relative to the full set of variables, we used a tour linked to a t-SNE layout using the R package `liminal` [@R-liminal]. This ensured that the final sample of clustered customers were also far apart in the high-dimensional space. (See the supplementary materials for further details.)<!--Few of these customers have similar distribution across *moy* and some are similar in their *hod* distribution. -->

>>>>

The proposed methods could be extended when two granularities of interest interact and we want to group subjects based on the interaction of the two granularities. Consider a group that has a different weekday and weekend behavior in the summer but not in the winter. This type of combined behavior across granularities can be discovered by evaluating the distribution across combinations of categories for different interacting granularities (weekend/weekday and month-of-year in this example). As a result, in this scenario, we analyze a combination of categories generated from different distributions. Display of design and related results can be found in [supplementary material](https://github.com/Sayani07/paper-gracsR/blob/master/paper-supplementary.pdf).

@@

All of the approaches were fitted to each data design and to each combination of the considered parameters considered in Table \ref{tab:range-parameter}. The formed clusters have to match the design, be well separated, and have minimal intra-cluster variation. <!--It is possible to study these desired clustering traits visually in a more comprehensive way than just looking at index values.--> MDS and parallel coordinate graphs are used to demonstrate the findings, as well as an index value plot to provide direction on the number of clusters. <!--JS-based approaches corresponding to NQT and RS are referred to as JS-NQT and JS-RS respectively.--> In the following plots, results for JS-NQT are reported, and results with JS-RS or wpd-based distances are in the [supplementary material](https://github.com/Sayani07/paper-gracsR/blob/master/paper-supplementary.pdf).

@@

Clustering with the new distances is illustrated on the smart meter energy usage for a sample of customers from @smart-meter. The full data contains half-hourly general supply in kWh for 13,735 customers, resulting in 344,518,791 observations in total.<!--In most cases, electricity data is expected to have multiple seasonal patterns like daily, weekly or annual. We do not learn about these repetitive behaviors from the linear view because too many measurements all squeezed in that representation. Hence we transition into looking at cyclic granularities, that can potentially provide more insight on their repetitive behavior.--> The raw data for these consumers is of unequal length, with varying starting and end dates. Additionally, there were missing values in many series. (The [supplementary material](https://github.com/Sayani07/paper-gracsR/blob/master/paper-supplementary.pdf) contains details from checking for systematic missingness.) Because our proposed methods evaluate probability distributions rather than raw data, these data issues are not problematic, unless there is any systematic structure related to granularities.

@@

It is common to filter data prior to fitting a supervised classification model using instance selection [@olvera2010review] which removes observations that might impede the model building. For clustering, this is analogous to identifying and removing nuisance observations. Prototype selection is more severe than instance selection, because only a handful of cases is selected. @Cutler1994-hm proposed a method called archetypal analysis which has inspired this approach but the procedure we have used follows @Fan2021-bq. First, dimension reduction such as t-SNE, MDS or PCA is used to project the data into a 2D space. Second, a few "anchor" customers far apart in 2D space are selected. Additional close neighbors to the anchors are selected. To check the selections relative to the full set of variables, we used a tour linked to a t-SNE layout using the R package `liminal` [@R-liminal]. This ensured that the final sample of clustered customers were also far apart in the high-dimensional space. (See the [supplementary materials](https://github.com/Sayani07/paper-gracsR/blob/master/paper-supplementary.pdf) for further details.)<!--Few of these customers have similar distribution across *moy* and some are similar in their *hod* distribution. -->

### 2022-02-28 Rmd/05-conclusion.Rmd
fine tuning all comments and cross checking with thesis all done

>>>>

Another long-term goal would be to develop a comparable framework with a spatial component for analysis. Electricity data is commonly found in a spatio-temporal format with some type of area code, such as a postcode or administrative unit. Assuming that households in certain residential neighborhoods have similar energy usage patterns, future research could look into clustering and visualization methods for analyzing probability distributions of spatio-temporal data, perhaps through some variation of the glyph maps.
